Millions of Type Errors

### 2023-09-09 programming23/paper.tex
cleanup old figs

\begin{table}[t]
  %% Looks the same when restricted to modules
  \begin{tabular}{l@{~~}l}
    \begin{tabular}[t]{lr}
      \code{SyntaxError} & 60.79\% \\
      \code{UnknownSymbol} & 27.34\% \\
      \code{UnknownProperty} & 4.08\% \\
      \code{TypeMismatch} & 1.92\% \\
      \code{UnknownRequire} & 1.40\% \\
      \code{CountMismatch} & 1.08\% \\
      \code{OptionalValueAccess} & 0.99\% \\
      \code{GenericError} & 0.93\% \\
      \code{CannotExtendTable} & 0.21\% \\
      \code{UnknownPropButFoundLikeProp} & 0.20\% \\
      \code{FunctionExitsWithoutReturning} & 0.16\% \\
      \code{ExtraInformation} & 0.14\% \\
      \code{CannotInferBinaryOperation} & 0.14\% \\
      \code{MissingProperties} & 0.13\% \\
      \code{ModuleHasCyclicDependency} & 0.10\% \\
      \code{IllegalRequire} & 0.09\% \\
      \code{NotATable} & 0.09\% \\
      \code{MissingUnionProperty} & 0.08\%
    \end{tabular} &
    \begin{tabular}[t]{lr}
      \code{CannotCallNonFunction} & 0.07\% \\
      \code{FunctionDoesNotTakeSelf} & 0.03\% \\
      \code{IncorrectGenericParameterCount} & 0.01\% \\
      \code{OnlyTablesCanHaveMethods} & 0.00\% \\
      \code{TypesAreUnrelated} & 0.00\% \\
      \code{DuplicateTypeDefinition} & 0.00\% \\
      \code{OccursCheckFailed} & 0.00\% \\
      \code{TypePackMismatch} & 0.00\% \\
      \code{FunctionRequiresSelf} & 0.00\% \\
      \code{DuplicateGenericParameter} & 0.00\% \\
      \code{InternalError} & 0.00\% \\
      \code{CodeTooComplex} & 0.00\% \\
      \code{DeprecatedApiUsed} & 0.00\% \\
      \code{NormalizationTooComplex} & 0.00\% \\
      \code{SwappedGenericTypeParameter} & 0.00\% \\
      \code{DynamicPropertyLookupOnClassesUnsafe} & 0.00\% \\
      \code{UnificationTooComplex} & 0.00\% \\
    \end{tabular}
  \end{tabular}
  \caption{Type error popularity}
  \label{t:type-error-count}
\end{table}

@@

\begin{table}[t]

  \caption{Type error survival}
  \label{t:type-error-survival}
\end{table}

@@

% TODO .. somewhere ... out/error-density-*
%   row->te nocheck :: 48378 ++ , 9440 ==, 45479 --
%   row->te nonstrict :: 19491 ++ , 9567 ==, 20121 --
%   row->te strict :: 733 ++ , 368 ==, 770 --
% 
% TODO ... somewhere ... te leq fs : use error-density

\begin{figure}[t]
  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row->te-density-diff.pdf}
  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row->te-density-diff.pdf}
  \includegraphics[width=\columnwidth]{img/error-count-strict-row->te-density-diff.pdf}

@@

Sixth, a mystery. May be a bug in Luau.
May be DM insensitivity in strict mode
(if DM has top type that would lead to errors.)
Context: each of the 3 modes runs 2 kinds of type analysis. One is the regular
type analysis. The other is Forced Strict (FS), which converts all dependencies
to strict mode. The reason for FS is that it sometimes gives better
autocomplete suggestions etc.
You'd expect that the number of type errors is always <= num FS errors. But we
found several cases where this is false!
The plot here "stacks" every strict session into bars. X-axis is a record index
within a session (bars are higher on the left because we have lots more short
sessions than long ones). Y-axis counts the number of records where num TE <= num FS
is true (blue) and false (yellow).

TODO clarify semantic x-axis order

%% FS has: 1 all dependencies in strict, 2 data-model awareness (DM turns into
%% type graph, ... initial DM, all properties go into a table type; DM -> folder
%% -> model -> part .. get dot-driven autocomplete) in strict, DM might have type
%% Any might have type Instance, the latter would mean more errors than FS

\begin{figure}[t]
  \includegraphics[width=0.8\columnwidth]{img/example-tefs.png}
  \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
  \label{f:tefs}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=0.8\columnwidth]{img/example-compass.png}
  \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
  \label{f:tefs}
\end{figure}

@@

\begin{figure}[h]
  \begin{tabular}{llrrr}
    Mode & E & Add & Keep & Remove \\\midrule
    \mnocheck{}   & \code{te} & - & - & - \\
                  & \code{fs} & - & - & - \\
    \mnonstrict{} & \code{te} & 16 & 18 & 21 \\
                  & \code{fs} & 19 & 18 & 18 \\
    \mstrict{}    & \code{te} & - & - & - \\
                  & \code{fs} & - & - & - \\
  \end{tabular}

  \caption{Does mode influence the change in total error count?}
  \label{f:edit-deltas}
\end{figure}

@@

\Cref{f:ex-session-strict} begins in strict mode
but jumps to nocheck mode as well.
The jumps sometimes coincide with module switches (vertical line),
sometimes not.
The green line counts type errors.in the codebase.

\begin{figure}[t]
  \includegraphics[width=0.7\columnwidth]{img/example-session-strict.pdf}
  \caption{Example strict session}
  \label{f:ex-session-strict}
\end{figure}

\Cref{f:ex-session-nonstrict} presents the three sessions from \mnonstrict{} mode
that have the greatest number of events (not necessarily the longest time spans).
There are no module or mode switches.
There are few type errors.
One high point shows 10 errors, but these errors are gone (probably fixed) at
the next event.

\begin{figure}[t]
  \includegraphics[width=0.7\columnwidth]{img/example-session-nonstrict.pdf}
  \caption{Three most populated nonstrict sessions}
  \label{f:ex-session-nonstrict}
\end{figure}

\Cref{f:ex-session-nocheck} shows a roller coaster from \mnocheck{} mode.
Several hundred ForcedStrict type errors appear and disappear between events.
There are lots of module switches too \ldots but that makes no difference
for these project-wide error counts.

\begin{figure}[t]
  \includegraphics[width=0.7\columnwidth]{img/example-session-nocheck.pdf}
  \caption{Example nocheck session}
  \label{f:ex-session-nocheck}
\end{figure}

\subsection{Aggregate Sessions}

\begin{figure}[t]
  \includegraphics[width=0.8\columnwidth]{img/fit-nocheck-Force-3.pdf}
  \includegraphics[width=0.8\columnwidth]{img/fit-nonstrict-TypeE-3.pdf}
  \includegraphics[width=0.8\columnwidth]{img/fit-strict-TypeE-3.pdf}
  \caption{Best-fit curves for errors in each mode (\code{fs} for \mnocheck{}, \code{te} for the others)}
  \label{f:scribbles}
\end{figure}

\Cref{f:scribbles} kill this.

@@

Greenman was supported by
grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.

>>>>

\begin{table}[t]
  \caption{Type error survival}
  \label{t:type-error-survival}

\end{table}

@@

\begin{table}[t]
  %% NOTE: looks the same when restricted to modules
  TODO recompute split by mode : te-editrange-ss-nocheck.rktd etc

  \begin{tabular}{l@{~~}l}
    \begin{tabular}[t]{lr}
      \code{SyntaxError} & 60.79\% \\
      \code{UnknownSymbol} & 27.34\% \\
      \code{UnknownProperty} & 4.08\% \\
      \code{TypeMismatch} & 1.92\% \\
      \code{UnknownRequire} & 1.40\% \\
      \code{CountMismatch} & 1.08\% \\
      \code{OptionalValueAccess} & 0.99\% \\
      \code{GenericError} & 0.93\% \\
      \code{CannotExtendTable} & 0.21\% \\
      \code{UnknownPropButFoundLikeProp} & 0.20\% \\
      \code{FunctionExitsWithoutReturning} & 0.16\% \\
      \code{ExtraInformation} & 0.14\% \\
      \code{CannotInferBinaryOperation} & 0.14\% \\
      \code{MissingProperties} & 0.13\% \\
      \code{ModuleHasCyclicDependency} & 0.10\% \\
      \code{IllegalRequire} & 0.09\% \\
      \code{NotATable} & 0.09\% \\
      \code{MissingUnionProperty} & 0.08\%
    \end{tabular} &
    \begin{tabular}[t]{lr}
      \code{CannotCallNonFunction} & 0.07\% \\
      \code{FunctionDoesNotTakeSelf} & 0.03\% \\
      \code{IncorrectGenericParameterCount} & 0.01\% \\
      \code{OnlyTablesCanHaveMethods} & 0.00\% \\
      \code{TypesAreUnrelated} & 0.00\% \\
      \code{DuplicateTypeDefinition} & 0.00\% \\
      \code{OccursCheckFailed} & 0.00\% \\
      \code{TypePackMismatch} & 0.00\% \\
      \code{FunctionRequiresSelf} & 0.00\% \\
      \code{DuplicateGenericParameter} & 0.00\% \\
      \code{InternalError} & 0.00\% \\
      \code{CodeTooComplex} & 0.00\% \\
      \code{DeprecatedApiUsed} & 0.00\% \\
      \code{NormalizationTooComplex} & 0.00\% \\
      \code{SwappedGenericTypeParameter} & 0.00\% \\
      \code{DynamicPropertyLookupOnClassesUnsafe} & 0.00\% \\
      \code{UnificationTooComplex} & 0.00\% \\
    \end{tabular}
  \end{tabular}
  \caption{Type error popularity}
  \label{t:type-error-count}
\end{table}

@@

\begin{figure}[t]\centering
TODO diff vs 1st, not previous

  \includegraphics[width=\columnwidth]{img/error-count-nocheck-row->te-density-diff.pdf}
  \includegraphics[width=\columnwidth]{img/error-count-nonstrict-row->te-density-diff.pdf}
  \includegraphics[width=\columnwidth]{img/error-count-strict-row->te-density-diff.pdf}

  TODO bars instead, percent only, don't care about counts

@@

Sixth, DM story.
FS has: 1 all dependencies in strict, 2 data-model awareness (DM turns into
type graph, ... initial DM, all properties go into a table type; DM -> folder
-> model -> part .. get dot-driven autocomplete) in strict, DM might have type
Any might have type Instance, the latter would mean more errors than FS

DM insensitivity in strict mode
(if DM has top type that would lead to errors.)
Context: each of the 3 modes runs 2 kinds of type analysis. One is the regular
type analysis. The other is Forced Strict (FS), which converts all dependencies
to strict mode. The reason for FS is that it sometimes gives better
autocomplete suggestions etc.
You'd expect that the number of type errors is always <= num FS errors. But we
found several cases where this is false!
The plot here "stacks" every strict session into bars. X-axis is a record index
within a session (bars are higher on the left because we have lots more short
sessions than long ones). Y-axis counts the number of records where num TE <= num FS
is true (blue) and false (yellow).

% TODO ... somewhere ... te leq fs : use error-density
%%

\begin{figure}[t]
  TODO use 2 bars? ask a few similar questions?
  \includegraphics[width=0.8\columnwidth]{img/example-tefs.png}
  \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
  \label{f:tefs}
\end{figure}

@@

\begin{figure}[t]
TODO smaller

  \includegraphics[width=0.5\columnwidth]{img/example-compass.png}
  \caption{Are there fewer type errors than forcedstrict errors? Expect true always, but that is not the case.}
  \label{f:tefs}
\end{figure}

@@

@@

@@

Thanks to Ben(jamin?) Chung for feedback on plots and data analysis advice.
Greenman was supported by
grant \href{https://nsf.gov/awardsearch/showAward?AWD_ID=2030859&HistoricalAwards=false}{2030859}
to the CRA for the \href{https://cifellows2020.org}{CIFellows} project.
